---
title: "AZLA_data_demo"
format: html
editor: visual
---

# Jung Mee Park

# jungmee.park@pima.gov

# 2025-9-16

```{r}
9+12
```
```{r}
print(29+1)
```
```{r}
print("hello world")
```

Read in data from this URL - https://raw.githubusercontent.com/jmp243/AZLA_2025_Data/refs/heads/main/fake_program_data.csv

```{r}
library(readr)
library(tidyverse)
```

```{r}
data <- read.csv("https://raw.githubusercontent.com/jmp243/AZLA_2025_Data/refs/heads/main/fake_program_data.csv")

```

# Data analysis

What type of data are we looking at?

```{r}
head(data)
```

```{r}
cls <- sapply(data, class)
cls <- as.data.frame(cls)
cls
```

A summary is not too helpful when we only have one variable that is an integer.

```{r}
summary(data)
```

```{r}
table(data$Branch.Location)
```

Look at other variables that way too.

```{r}
table(data$Target.Audience)
```

Create an reference number for each event

```{r}
data_with_ref <- data %>%
  mutate(Reference_Number = row_number())

```

Parse out the age groups 

```{r}

data2 <- data_with_ref %>%
  # split Target.Audience at "&" and make multiple rows
  separate_rows(Target.Audience, sep = " *& *")  %>%
  # optional: trim whitespace
  mutate(Target.Audience = trimws(Target.Audience))
```

Create a new column where age distinctions do not matter
```{r}
library(stringr)

data2_newAudience <- data2 %>%
  mutate(
    # extract everything before " (" (a space + "(") or just "("
    AudienceGroup = str_trim(
       str_replace(Target.Audience, "\\s*\\(.*", "")
    )
  )
```

look at the time of day when events were happening
```{r}
library(lubridate)
# Using rowwise() to parse time

data2_time <- data2_newAudience %>%
  rowwise() %>%
  mutate(
    # Process one row at a time
    time_parsed = {
      time_str <- Time
      if (is.na(time_str) || !is.character(time_str)) {
        NA
      } else if (grepl("AM|PM", time_str, ignore.case = TRUE)) {
        parse_date_time(time_str, orders = c("I:M p", "H:M p"))
      } else if (grepl(":", time_str, fixed = TRUE)) {
        if (length(gregexpr(":", time_str)[[1]]) == 2) {
          parse_date_time(time_str, orders = "H:M:S")
        } else {
          parse_date_time(time_str, orders = "H:M")
        }
      } else {
        NA
      }
    }
  ) %>%
  ungroup() %>%
  mutate(
    time_24h = format(time_parsed, "%H:%M:%S"),
    time_12h = format(time_parsed, "%I:%M:%S %p")
  )

```

group into hours
```{r}
# group pits incidents by the hour

# Group by hour
data2_time$hour_group <- floor_date(data2_time$time_parsed, 
                                               unit = "hour")
data2_time$hour_label <- format(data2_time$hour_group, "%H:00")

```

# fix dates
```{r}
data2_time$Date <- as.Date(data2_time$Date,format = "%m/%d/%Y")
```


# Visualizations
```{r}
library(plotly)
# Group by hour and count events
hourly_summary <- data2_time %>%
  select(hour_label, Branch.Location, Attendance, time_parsed, Date) %>%
  # mutate(hour_group = floor_date(time_parsed, unit = "hour")) %>%
  group_by(hour_label) %>%
  summarize(
    Total_Hourly_Count = n(),
    # other aggregate metrics as needed
  )

```
You can see which hours had the most events.

```{r}
hourly_location_date_summary <- data2_time %>%
  select(hour_label, hour_group, Attendance, AudienceGroup, Branch.Location, time_parsed, Date) %>%
  # Group by date as well as hour and location
  group_by(Date, hour_label, AudienceGroup, Branch.Location, Attendance) %>%
  summarize(
    count = n(),
    .groups = "drop"
  ) %>%
  # Create a wide format with Location combinations as columns
  pivot_wider(
    id_cols = c(Date, hour_label),  # Keep Date2 as an identifier column
    names_from = AudienceGroup,
    values_from = Attendance,
    values_fill = 0
  )

```

creating a visualization can be helpful.

## group data by attendance
```{r}
# Load required libraries
library(dplyr)
library(plotly)

# Aggregate attendance by hour_group and AudienceGroup
plot_data <- data2_time %>%
  mutate(hour_group = as.factor(hour_label)) %>%
  group_by(hour_label, AudienceGroup) %>%
  summarise(Total_Attendance = sum(Attendance, na.rm = TRUE), .groups = "drop")

# Create interactive bar chart
fig <- plot_ly(
  data = plot_data,
  x = ~hour_label,
  y = ~Total_Attendance,
  color = ~AudienceGroup,
  type = "bar"
)

# Customize layout
fig <- fig %>%
  layout(
    barmode = "stack",   # or "group" if you want side-by-side
    xaxis = list(title = "Hour Label"),
    yaxis = list(title = "Total Attendance"),
    legend = list(title = list(text = "Audience Group"))
  )

fig
```

### change x-axis to dates
```{r}
# Load required libraries
library(dplyr)
library(plotly)

# Aggregate attendance by hour_group and AudienceGroup
data2_time <- data2_time %>% 
  filter(Date >= 2025)

plot_data <- data2_time %>%
  mutate(hour_label = as.factor(hour_label)) %>%
  group_by(Date, AudienceGroup) %>%
  summarise(Total_Attendance = sum(Attendance, na.rm = TRUE), .groups = "drop")

# Create interactive bar chart
fig_date <- plot_ly(
  data = plot_data,
  x = ~Date,
  y = ~Total_Attendance,
  color = ~AudienceGroup,
  type = "bar"
)

# Customize layout
fig_date <- fig_date %>%
  layout(
    barmode = "stack",   # or "group" if you want side-by-side
    xaxis = list(title = "Date"),
    yaxis = list(title = "Total Attendance"),
    legend = list(title = list(text = "Audience Group"))
  )

fig_date
```


## by branch type data graph
```{r}
# Aggregate attendance by hour_group and AudienceGroup
branch_data <- data2_time %>%
  group_by(hour_label, Branch.Location) %>%
  summarise(Total_Attendance = sum(Attendance, na.rm = TRUE), .groups = "drop")

# Create interactive bar chart
fig2 <- plot_ly(
  data = branch_data,
  x = ~hour_label,
  y = ~Total_Attendance,
  color = ~Branch.Location,
  type = "bar"
)

# Customize layout
fig2 <- fig2 %>%
  layout(
    barmode = "stack",   # or "group" if you want side-by-side
    xaxis = list(title = "Hour Label"),
    yaxis = list(title = "Total Attendance"),
    legend = list(title = list(text = "Branch Location"))
  )

fig2
```

## add a date slider
```{r}
# Prepare data: make sure Date is in Date format
plot_data <- data2_time %>%
  mutate(
    Date = as.Date(Date),
    hour_group = as.factor(hour_label)
  ) %>%
  group_by(Date, hour_label, AudienceGroup) %>%
  summarise(Total_Attendance = sum(Attendance, na.rm = TRUE), .groups = "drop")

# Build interactive bar chart
fig <- plot_ly(
  data = plot_data,
  x = ~Date,
  y = ~Total_Attendance,
  color = ~AudienceGroup,
  type = "bar",
  hoverinfo = "text",
  text = ~paste("Date:", Date,
                "<br>Hour:", hour_label,
                "<br>Audience:", AudienceGroup,
                "<br>Attendance:", Total_Attendance)
)

# Layout with slider
fig <- fig %>%
  layout(
    xaxis = list(
      title = "Date",
      rangeselector = list(
        buttons = list(
          list(count = 7, label = "1w", step = "day", stepmode = "backward"),
          list(count = 1, label = "1m", step = "month", stepmode = "backward"),
          list(step = "all")
        )
      ),
      rangeslider = list(visible = TRUE),
      type = "date"
    ),
    yaxis = list(title = "Total Attendance"),
    barmode = "stack",
    legend = list(title = list(text = "Audience Group"))
  )

fig
```

## create a shiny app
```{r}
library(shiny)
library(dplyr)
library(plotly)


```

